Locality and Effect Reflection

### 2025-09-15 paper.tex
Shorten the sums subsection by removing Half_term

\lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}. Terms in
the free algebras of the sum of two effects
(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
recursive datatype that is built from the terms of one of the effects
and the operations of the other~\cite{hyland2006combining}. We can define
this type and associated isomorphism as:
\begin{lstlisting}[style=oxcaml]
  module Half_term (L : Op) (R : Op) : sig
    type 'a t =
      | Return : 'a @@ global -> 'a t
      | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
  end

  module Half_reflection (L : Op) (R : Op) : sig
    val reify :
      (Handler(L).t @ local -> Handler(R).t @ local -> 'a)
      -> Handler(L).t @ local -> 'a Half_term(L)(R).t

    val reflect :
      (Handler(L).t @ local -> 'a Half_term(L)(R).t)
      -> Handler(L).t @ local -> Handler(R).t @ local -> 'a
  end
\end{lstlisting}

Note that the key difference between
\lstinline[style=oxcaml]{Half_term(L)(R).t} and
\lstinline[style=oxcaml]{Term(R).t} is that continuing the computation
after an operation requires passing in a local
\lstinline[style=oxcaml]{Handler(L).t} again.

Now we can use this machinery to conveniently build a handler of
\lstinline[style=oxcaml]{Gen} that also forwards operations from
\lstinline[style=oxcaml]{Await}:
\begin{lstlisting}[style=oxcaml]
  module Await_gen_eff = Half_reflection(Await)(Gen)
  open Half_term(Await)(Gen)

@@

  let handle_gen_in_await f ah =
      let rec loop = function
      | Return () -> Finished
      | Op(Gen i, k) -> More(i, fun ah -> loop (k () ah))
    in
    loop (Await_gen_eff.reify f ah)
\end{lstlisting}
where \lstinline[style=oxcaml]{handle_gen_in_await} has the type:
\begin{lstlisting}[style=oxcaml]
  val handle_gen_in_await :
    (Handler(Await).t @ local -> Handler(Gen).t @ local -> unit)
    -> Handler(Await).t @ local
    -> aints
\end{lstlisting}

The literature on lexical effect handlers often ignores this issue. For
example, Xie et al.~\cite{xie2020effect} proposed a system which only
supported directly nesting effect handlers and had no mechanism for
writing a composable handler for generators that could be moved between
different underlying asynchronous I/O handlers.

This is because composition requiring forwarding is much less common
than nesting and it fundamentally requires each use of the
\lstinline[style=oxcaml]{Await} operation to perform work proportional
to the number of handlers it is forwarded through -- often referred to
as a ``linear search''. There has been a tendency to treat this linear
search as some form of deficiency of algebraic effect handlers, but it
is simply a classic trade-off between the extra expressivity allowed by
an indirection and the runtime cost of traversing that indirection. With
our effect reflection API we are able to use nesting for the cases that
it supports -- avoiding the cost of the indirection -- and fall back to
forwarding for the more general cases.

>>>>

We can now write our desired handler for the
\lstinline[style=oxcaml]{Gen} effect by instead writing a handler for
\lstinline[style=oxcaml]{Sum(Await)(Gen)} that forwards all the
\lstinline[style=oxcaml]{Await} operations on to another handler.

\begin{lstlisting}[style=oxcaml]
  module Eff = Reflection(Sum(Await)(Gen))
  module Proj = Project(Await)(Gen)

@@

  let handle_gen_in_await f ah =
    let rec loop ah = function
      | Return () -> Finished
      | Op(Left op, k) -> loop ah (k (perform ah op))
      | Op(Right(Gen i), k) -> More(i, fun ah -> loop ah (k ()))
    in
    loop ah (Eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))
\end{lstlisting}

The need for effect forwarding is a general short-coming with lexical
handlers that applies to all such systems. Examples like this
fundamentally require that performing an \lstinline[style=oxcaml]{Await}
operation to does work proportional to the number of handlers it is
forwarded through. Lexical handlers do a constant amount of work to
perform operations and so cannot handle such examples
directly. Non-lexical effect handlers *always* do work proportional to
the number of nested handlers, allowing them to support such examples
directly but paying a higher cost for the cases that do not need
forwarding.
